Visualizing Independence Using Extended Association Plots

### 2005-08-03 trunk/vcd/paper/JCGS.Rnw
finished Sections 3.1 and 3.2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@427 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

colors and color spaces

@@

Colors are commonly used to enhance mosaic and association plots. To integrate
a visualization of the Pearson residuals $\{r_{ij}\}$ into the mosaic display---which
in its `raw' version only visualizes the observed frequencies $\{n_{ij}\}$---\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}. 
In this section, we first briefly review the \cite{vcd:Friendly:1994} shading, 
before we suggest different colors and a combination of visualization and significance
testing.

@@

A first extension by
\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table, that is, from
the expected frequencies under independence. This approach does not only work in 2-way
tables but also in log-linear models fitted to multi-way tables. In this
extension, positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide
substantial improvement of the original mosaic plots and enhanced them
from a plot for contingency tables to a visualization technique for
log-linear models (and therefore also for certain independence problems).
However, it has two major drawbacks: First,
the significance level for the hard-coded critical values 2 and 4 is
usually unknown. Second, the colors for (light) blue and red
used in the Michael Friendly's \textsf{SAS} implementation and also in the base implementation
in \textsf{R} are not device-independent and
do not provide homogeneous saturations over different colors
and copier proofness. We suggest some ways to overcome these drawbacks.

The implementation of Friendly mosaic plots in \textsf{SAS}
uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
with hues for blue and red, full saturation, and varying luminance
for lighter colors. The current \textsf{R} implementation is
based on colors in the HSV (Hue--Saturation--Value) space where
decreasing the saturation from 1 towards 0
%% ---given the hue for blue or red and full saturation---
generates very similar colors
as increasing the luminance for HLS colors from 0.5 towards 1.
As an alternative to those HLS and HSV colors, the device-independent
HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
used which also provides homogeneous saturations over different colors and copier
proofness. The resulting association plot (using critical values for
$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
there is no evidence for rejecting the hypothesis of independence
in this table. Thus, there seems to be neither an advantage for
the home team nor for the away team.

@@

currently: chosen based on HSV or HLS color spaces,
hue (red and blue) codes sign of residuals, and saturation the absolute size
in three steps, value currently unused (always set to full value)

disadvantages:
  - perceptually not uniform
    because the three perceptual dimensions (hue, lightness, saturation)
    are not properly mapped to the three dimensions of the color space
    and hence are confounded \citep{vcd:Brewer:1999}    
    in particular, saturation is not uniform across different hues.

  - can lead to color-caused optical illusions in statistical graphs
    \citep{vcd:Cleveland+McGill:1983}

  - flashy fully saturated HSV colors are good for drawing attention to a plot,
    but hard to look at for a longer time, hence mosaic displays using these 
    colors are harder to interpret

  - white is not very suitable as a neutral color, grey conveys neutrality or 
    uninterestingness much better

Alternatives:
  - \cite{vcd:Munsell:1905} introduced color notation for balanced colors
  - Commission Internationale de l'\'Eclairage (CIE) introduced two perceptually
    based color spaces CIELAB and CIELUV where the latter is preferred for 
    emissive color technologies such as computer displays.
  - \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} is an online tool
    for selcting different sets of colors, based on similar ideas. It includes
    qualitative, sequential and diverging palettes.
  - \cite{vcd:Ihaka:2003} discusses how CIELUV colors can be used for choosing
    colors for statistical graphics such as barplots. He recommends to use
    polar coordinates in the CIELUV space---called HCL space, using hue, chroma
    and luminance---and suggests a strategy for choosing qualitative schemes based
    on these.

We discuss how similar ideas can be used for deriving a diverging HCL scheme that 
is suitable for visualizing residuals in association and mosaic displays.

HCL
  - irregular double cone, i.e., chroma/luminance plane looks rather different
    for different hues
  - choose two hues that look similar in the chroma/luminance plane to code
    the sign of the residuals. Also use red and blue at full chroma for large
    residuals and use a light gray (instead of a white at the top of the luminance
    scale) as the neutral color for small residuals, interpolate between.
  - for non-significant results, clearly reduce the amount of color, i.e. the maximum
    chroma used.

Advantages
  - based on a perceptually uniform color space
  - device independent

@@

## plot 2 slides from HCL space
plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
for(i in 1:(nc-1)) {
  rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
}
for(i in 1:(nc-1)) {
  rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
}
axis(2, at = c(50, 75, 100), labels = c(50, 75, 100))
axis(4, at = c(50, 75, 100), labels = c(50, 75, 100))
axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
mtext("chroma", side = 3, at = 0, line = 2)
mtext("luminance", side = 2, at = 75, line = 2)
mtext("luminance", side = 4, at = 75, line = 2)
lines(c(-100, 100), c(50, 50))

@@

lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
@
\caption{Extended shading in HSV and HCL space.}
\label{fig:shadingHSVHCL}
\end{center}
\end{figure}

@@

>>>>

colors and color spaces
    because the three perceptual dimensions (hue, lightness, saturation)
    are not properly mapped to the three dimensions of the color space
    and hence are confounded \citep{vcd:Brewer:1999}

@@

Colors are commonly used to enhance mosaic and association plots. To integrate
a visualization of the Pearson residuals $\{r_{ij}\}$ into the mosaic display---which
in its `raw' version only visualizes the observed frequencies $\{n_{ij}\}$---\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}. 
In this section, we first briefly review the \cite{vcd:Friendly:1994} shading, 
before we suggest different colors and a combination of visualization and significance
testing to extend these residual-based shadings.

@@

The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide a
substantial improvement of the original mosaic plots and enhanced them
from a plot for contingency tables to a visualization technique for
log-linear models---and thus also for independence problems including
2-way tables as the simplest special case.

The idea is to use a color coding for the mosaic tiles that visualizes the
sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
residuals ($|r_{ij}| < 2$) are shaded white. Cells with medium sized residuals
($2 \le |r_{ij}| < 4$) are shaded in light blue and light red for positive negative
residuals respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
with a fully saturated blue and red respectively. Mosaic plots enhanced by this 
shading can thus also bring out departures from independence (or other log-liner 
models in multi-way tables) graphically and visualize patterns of dependence.
The heuristic for choosing the cut offs $2$ and $4$ is that the Pearson residuals
are approximately standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha = 0.05$ and $\alpha = 0.0001$
level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

In addition to the shading of the rectangles themselves, the Friendly shading also
encompasses a choice of line type and line color of the borders of the rectangles
with similar ideas as described above. As both mosaic and association plots are
area-proportional visualization techniques, we focus on area shadings and always 
use solid black borders throughout this paper, but the extensions suggested in the
following could also be applied to control line type and color.

@@

The way the (light) blue and red colors are chosen differs somewhat between
various implementations of Friendly mosaic plots: In his original 
\proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
uses colors from a palette based on HLS (Hue--Luminance--Saturation) color space.
The implementation in the \proglang{R} system for statistical computing
\citep[\url{http://www.R-project.org/}]{vcd:R:2005} employs colors from
HSV (Hue--Saturation--Value) space. The latter is a very common implementation of
colors in many computer packages and makes the generation of the Friendly shading
very simple.

The HSV space originally looks like a regular cone with black at its
peak (zero value) and full color wheels for different saturations at the other end,
centered around a white (full value). Type and amount of color are controlled by hue
and saturation respectively. Typically, polar coordinates that all range in the 
unit interval are used in this space, such that it looks like a regular cylinder.
For generating colors in the Friendly shading, the following strategy is used:
The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
$h = 2/3$ (blue hue) for positive residuals. The absolute size of the residuals
is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$ respectively
for small/medium/large residuals respectively. The value is always fixed at $v = 1$.
This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
to slides from the HSV cylinder side to side. The plot shows the saturation/value plane
for the given hues $h = 0$ and $h = 2/3$. The significant color palette shows the
colors used for the Friendly shading when the residuals are increasing from left to right.
The set of non-significant colors will be explained in the next section.

Although this HSV-based shading is already very useful for enhancing mosaic and
association plots and although HSV is very commonly available implementation of
color spaces, HSV colors in general and the Friendly shading in particular have
a number of disadvantages. Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \citep{vcd:Brewer:1999,vcd:Ihaka:2003}.
Consequently, the the HSV dimensions are confounded, e.g., saturation
is not uniform across different hues. A fully saturated blue $(2/3, 1, 1)$ is
much darker as a fully saturated red $(0, 1, 1)$ or green $(1/3, 1, 1)$.
This makes it more difficult for the human eye to judge the size of shaded areas
and can therefore lead to color-caused optical illusions when used in statistical graphs
\citep{vcd:Cleveland+McGill:1983}. Furthermore, flashy fully saturated HSV colors are
good for drawing attention to a plot, but hard to look at for a longer time \citep{vcd:Ihaka:2003}
which makes HSV-shaded graphics harder to interpret. Finally, white is employed as
the neutral color for small residuals in the Friendly shading, but typically
grey is found to convey neutrality or uninterestingness much better than white does
\citep{vcd:Brewer:1999}.

Alternative ways to choose colors have been available for a long time, but have
been only slowly adopted for implementations of colors in computer packages in 
general and for shading in statistical graphs in particular. The idea of using
perceptually based colors that are `in harmony' go back to \cite{vcd:Munsell:1905} 
who introduced a color notation for balanced colors. Based on similar principles, Cynthia
Brewer and co-workers suggested different types of palettes (qualitative/sequential/diverging)
and provided the online tool \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} for
selecting an appropriate palette for a specific problem. Furthermore, the
Commission Internationale de l'\'Eclairage (CIE) introduced the two perceptually
based color spaces CIELAB and CIELUV where the latter is typically preferred for 
emissive color technologies such as computer displays. \cite{vcd:Ihaka:2003} discusses
how CIELUV colors can be used for choosing colors for statistical graphics such as barplots.
By taking polar coordinates in CIELUV space, it is called HCL (Hue--Chroma--Luminance) space
and qualitative palettes can easily be chosen by using a range of hues for fixed values
of chroma and luminance. Such colors are always balanced towards the same grey 
and thus do not have the problem of varying saturations as for the HSV colors.

Here, we discuss how similar ideas can be used for deriving a diverging HCL palettes that 
provide a suitable translation of the ideas from the Friendly shading to perceptually
uniform HCL colors. 
The HCL space looks like a double cone with black (zero luminance) at one end and white
at the other (full luminance). In its middle, there is a full color wheel for different
values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not completely
regular as the HSV space is and consequently its dimensions cannot be standardized to 
unit intervals: the hue ranges in $[0, 360]$ degrees and chroma and luminance both in $[0, 100]$
percent. But not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
of $c$ and $l$ vary across different hues $h$. But for the problem of constructing
a diverging palette, this problem can easily be overcome as we just need two different
hues (a `negative' and a `positive' hue) and hence choose two hues that correspond to 
similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
shows two such planes side to side for the hues $h = 0$ and $h = 260$. To obtain a 
sequence of colors with the same properties as the Friendly shading, the palette starts
at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
blue $(260, 100, 50)$, and uses linear interpolation in between. Instead of using white
$(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.

This diverging palette (see the `significant' colors in Figure~\ref{fig:shadingHSVHCL},
the `non-significant' colors are again discussed in the following section)
uses both chroma, i.e., the colorfulness, and luminance, i.e., 
the amount of grey, to code the absolute size of the quantity visualized, i.e., the
Pearson residuals when applied to the independence problem. By changing the neutral
color or by changing the maximum chroma respectively, this can be changed to using only
chroma or luminance for this purpose, but using both seems to be a very effective way
of visualization.

The advantages of using this HCL shading scheme instead of the HSV scheme is that
it is based on a perceptually based color space that provides uniform saturations
for different hues and is device inependent.

@@

## plot 2 slides from HCL space
plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
for(i in 1:(nc-1)) {
  rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
}
for(i in 1:(nc-1)) {
  rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
}
axis(2, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
axis(4, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
mtext("chroma", side = 3, at = 0, line = 2)
mtext("luminance", side = 2, at = 75, line = 2)
mtext("luminance", side = 4, at = 75, line = 2)
lines(c(-100, 100), c(50, 50))

@@

lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
@
\caption{Extended shading in HSV (upper) and HCL space (lower).}
\label{fig:shadingHSVHCL}
\end{center}
\end{figure}

@@

However, it has two major drawbacks: First,
the significance level for the hard-coded critical values 2 and 4 is
usually unknown.

  - for non-significant results, clearly reduce the amount of color, i.e. the maximum
    chroma used.
